\documentclass[defaultstyle,11pt,english]{thesis}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%         PACKAGES & CONFIG        %%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% Your thesis gets printed in color---you can use colors!  This package has
% something to do with enabling that?
\usepackage{color}

% Improved typography.
%\usepackage{csquotes}
\usepackage{microtype}

% Improved text alignment.
\usepackage{ragged2e}

% To insert figures.
\usepackage{graphicx}
\graphicspath{{chapter-2/figs}, {chapter-3/figs}, {chapter-4/doodles}, {chapter-5/figs}, {chapter-6}}
%
\usepackage{epstopdf}
% For sub-figure support.  I think the subcaption package may be better, but
% that doesn't work with revtex (the document class for Physical Review), so I
% use this one.
\usepackage[caption=false]{subfig}
% Honestly I forget why I included this package...
\usepackage[export]{adjustbox}
% You can use this command to put a box around LaTeX-generated figures.
%\usepackage{longfbox}

% Lists.
%\usepackage{enumitem}

% Proper equation rendering.
\usepackage{amsmath}
\usepackage{amssymb}
%\usepackage{bbold}
\usepackage{bm}
\usepackage{mathtools}
% Introduces \bra{} and \ket{} commands.  Never type \rangle again.
\usepackage{braket}

% For formatting quotes
\usepackage{multicol}

% Improved tables.
\usepackage{array}
\usepackage{booktabs}
%\usepackage{multirow}
\usepackage{dcolumn}
\usepackage{tabularx}
%\usepackage{makecell}
% Alignment in table columns.
\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\newcolumntype{R}[1]{>{\raggedleft\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
% Increase table spacing so they're actually legible.
\renewcommand{\arraystretch}{1.5}

% PDF hyper-references (i.e., links between pages).
\usepackage{hyperref}
% This makes hyper-references not hideous.  Without this, links are black text
% with an ugly red box around them.  If you're fine with that, do you even
% deserve to graduate?
\usepackage[dvipsnames]{xcolor}
\hypersetup{
  colorlinks,
  linkcolor={red!50!black},
  citecolor={blue!50!black},
  urlcolor={blue!80!black},
  pdftitle={Exotic Dynamics}
}

% HACK: Add a tiny bit of space atop all floats (figures and tables) to prevent
% them from overlapping with page numbers when positioned at the top of the
% page.  There ought to be a better way...
\makeatletter
\g@addto@macro\@floatboxreset{\vspace{2pt}}
\makeatother

% For example:
% This makes all the quote environments italicized.
\AtBeginEnvironment{quote}{\itshape}

\usepackage{tikz-cd}
\usepackage{soul}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%       BIBLIOGRAPHY SETUP        %%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% https://mirror.mwt.me/ctan/macros/latex/contrib/x/doc/biblatex.pdf
% https://mirror.mwt.me/ctan/macros/latex/contrib/biblatex-contrib/biblatex-phys/biblatex-phys.pdf
\usepackage[
  backend=bibtex,
  style=phys, % The APS citation style.
  articletitle=true, % References without article titles are so annoying.
  biblabel=brackets,
  pageranges=false, % Does anybody care about page ranges anymore?
  % backref=true, % Enables "Cited on p. XX" messages for all references.
  sorting=nyt,
  sortcites=true,
  sortcase=false,
  url=false, % Still printed for @online entries.
  eprint=false,
  doi=false, % Provide missing page numbers to suppress this everywhere
  date=year,
]{biblatex}
% Allows the bibliography to create DOI links.
\usepackage{doi}
% The name of your bibliography file.  Pretty sure you can import multiple.
\addbibresource{bibliography.bib}
\addbibresource{bob.bib} 
% Defines the format for back-references.
\DefineBibliographyStrings{english}{%
  backrefpage = {Cited on p.},
  backrefpages = {Cited on pp.},
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%             PREAMBLE            %%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% The title of your thesis.
\title{Quantum Dynamics with Exotic Symmetries}
% Uncomment this if you want to customize the display of your thesis' title on
% the title page only (such as inserting a line break).
% \titlePageTitle{
%   This is the Name\\
%   of my Thesis
% }

% Who even are you ?
\author{Charles Nicholas} % {First Middle}
  {Stahl} % {Last}

% How long have you already been in school?
\otherdegrees{
  A.B., Princeton University, 2018 \\
  M.Ast., Cambridge University, 2019
}

% What degree is this thesis gonna get you?
\sodef\spaceout{}{0pt plus 1fil}{.7em plus 1fil}{0pt}
\degree{\parbox{1.7in}{\makebox[\linewidth][l]{\spaceout{Doctor of Philosophy}}}} % {Long description}
  {Ph.D., Physics}  % {Short description}

% What department?
\dept{Dept. of} % {Designation}
  {Physics} % {Name}

% The chair of your thesis committee.
\advisor{Professor} % {Title}
  {Rahul Nandkishore}  % {Name}
% Other members of the thesis committee.
\reader{Michael Hermele}
\readerThree{Andrew Lucas}
\readerFour{Leo Radzihovsky}
\readerFive{Markus Pflaum}

% Your abstract will probably be more than one page.
\abstract{ \OnePageChapter  % because it is very short
  Symmetries have a long history of importance in physics, especially in the study of dynamics and in quantum many-body systems. From Noether's theorem describing conserved quantities and Landau's program of spontaneous symmetry breaking, to higher-form symmetries in quantum memories and multipole symmetries in Hilbert-space fragmentation, symmetries have provided routes to constraining quantum dynamics. In particular, they can often lead to physical systems being able to store information.

  In this thesis we present some ways that exotic and modern notions of symmetry can lead to interesting nontrivial quantum dynamics. In the first half, consisting of three chapters, we present a series of projects on higher-form symmetry-protected self-correctiong quantum memories and single-shot quantum error correction. We generalize symmetry-protected quantum memories to paramagnetic and Walker-Wang bulks, along with topologically ordered bulks. With the topologically ordered bulk, the symmetry need only be enforced on a boundary. We then construct a new presentation of a single-shot quantum memory and highlight its close connection to symmetry-protected quantum memories.
  
  The second half of this thesis consists of two chapters, in which we address two problems related to Hilbert-space shattering and exotic symmetries. First, we study the dynamics of dipole-conserving systems in the hydrodynamics regime. The existence of this regime assumes that the shattering does not play a role, but nevertheless the dipole symmetries have a nontrivial affect on the dispersion relations. Second, we construct a model of shattering that is robust to arbitrary local symmetric perturbations, or to generic perturbations within a prethermal timescale. This model relies on a higher-form symmetry for its robustness.
}

\dedication[]{ % NEVER use \OnePageChapter here.
  % Use \begin{center} ... \end{center} to center this content.
  \vspace{2cm}
  For $\mathcal{H}$aley,
  
  \vspace{2cm}
  My best friend, always
}

\acknowledgements{ \OnePageChapter 
  Having Rahul Nandkishore as my advisor has been such a great privilege. From my early days as a grad student through my process of gaining some independence as a researcher, he has always given me exactly the amount of direction that I need, while also giving me the freedom to explore the projects that I want to. His research and career advice has always landed, has always been useful and correct, and has made it clear he wants me to succeed, not just as his student but as my own researcher. I also want to thank Andy Lucas and Mike Hermele for many interesting and encouraging conversations, and, along with Markus Pflaum and Leo Radzihovsky, for serving on my committee and agreeing to read my thesis.
  
  I'm grateful for all the entertainment and education I've recieved from the students and postdocs in the CTQM. My officemates Mert Okyay and Matteo Wilczak have always kindly listened to whatever I'm working on or talking about. The postdocs Ollie Hart, David Stephen, and Aaron Friedman have generously taken me in from wandering the halls looking for someone to bother. And all of the other grad students, especially Kenny Higginbotham, Marvin Qi, Evan Wickenden, Yifan Hong, and Jack Farrell, have been the sources of countless chats, ideas, and explanations.
  
  I would not be where I am today without my family. For teaching me, believing in me, and always pushing me to be my best, I am forever grateful.
  
  And of course, to Haley, thank you for all of your love and support. For being goofy with me and for being serious with me. For traveling with me and for staying put with me. Thank you for listening to and being interested in my long rants about quantum dynamics and materials and memory. This thesis would not exist without you.
}

\introductoryPage{}

\emptyLoT  % use this if there is no List of Tables


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%          From QF Model          %%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\usepackage[inline,shortlabels]{enumitem}
\usepackage[export]{adjustbox}
\usepackage{dsfont}
\usepackage{pifont}
\usepackage{stackengine}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%         CUSTOM COMMANDS         %%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\input macros.tex
\input chapter-6/tikzpics.tex

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%            DOCUMENT             %%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

% FONT: ecrm1095
% TEXTWIDTH: 469.75499pt = 6.52437486112in
% BODY FONT SIZE: 10.95pt
% FIGURE FONT SIZE: ^ same

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%            CHAPTERS             %%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\input chapter-1/chapter-1.tex
\input chapter-2/1formWalkerWang.tex
\input chapter-3/defectmem.tex
\input chapter-4/main.tex
\input chapter-5/multipoleSuperfluids.tex
\input chapter-6/condensed.tex
\input chapter-7/conclusion.tex
\nocite{Carroll2002Alice}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%          BIBLIOGRAPHY           %%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{singlespacing}
\printbibliography[
  % Title of this section.
  title={References},
  % This line includes the bibliography in the table of contents, which is
  % required by the graduate school.
  heading=bibintoc
]
\end{singlespacing}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%          APPENDICES           %%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\appendix
\input chapter-4/appendix.tex
\input chapter-5/appendix.tex
\input chapter-6/appendix.tex

\end{document}